---
title: "GENIE Analysis example"
output:
  html_document:
    df_print: paged
---

This is an example notebook of how to interact with the public GENIE data.
This notebook assumes that you already have a synapse account and you have
accepted to the the GENIE data conditions for use.

## Using `synapser`

`synapser` is the `R` Synapse client. It allows you to interact with files in Synapse in an R session.
The installation instructions can be found [here](https://r-docs.synapse.org/).

Once you have installed `synapser`, you will want to set up your [R Synapse credentials](https://r-docs.synapse.org/articles/manageSynapseCredentials.html)

```{r}
library(synapser)
synLogin()
```


## Getting the data

Once you have installed `synapser`, you are ready to interact with the data!
Please head to the [GENIE site](synapse.org/genie) and determine which release
you would like to interact with.

Everything in Synapse has a Synapse Id.  If you want to download the sample file
for release 8.0, you will click on the "Files" tab, then navigate through "Data Releases"
and click [Release 8.0-public](https://www.synapse.org/#!Synapse:syn22228642).
You will see that the sample file has Synapse Id: syn22228695

```{r}
sample_ent = synGet("syn22228695")
print(sample_ent$path)
```


## Your analysis

As a simple example, I want to see the distribution of AGE_AT_SEQ_REPORT for GENIE samples.

```{r}
sampledf = read.csv(sample_ent$path, sep = "\t", comment.char = "#")

hist(as.numeric(sampledf$AGE_AT_SEQ_REPORT),
     main = "Distribution of AGE_AT_SEQ_REPORT",
     xlab = "Age",
     ylab = "Number of Samples")
```

I also want to determine the number of samples per oncotree code


```{r}
oncotree_codes = table(sampledf$ONCOTREE_CODE)
write.csv(oncotree_codes, "oncotree_code_dist.csv", row.names = F)
```
